---
title: "Bacteria Abundance Heatmap and Dendogram Analysis"
author: "Martin Rühle"
date: "24/10/24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Main objective of the script
This section provides an overview of the analysis. The goal is to create a heatmap with a dendrogram of relative bacterial abundances grouped by metadata categories. The script will preprocess the data, filter it based on variance, and visualize the data using a heatmap with annotated categories like race and term status.

```{r}
# Load necessary libraries
library(pheatmap)
library(dplyr)
library(RColorBrewer) # For custom color palettes
```

## Load and Preprocess Data
In this section, the necessary datasets are loaded into R. The `metadata` file contains information about each sample, such as race and whether it was a full-term pregnancy, while the `rel_abun` file contains the relative abundances of various bacterial species. The two files are loaded as data frames and will later be combined for analysis.
```{r}
# Read the metadata and relative abundance files
metadata <- read.csv("C:/Users/User/Documents/DREAM/StudyFiles/Training/Training/metadata/metadata.csv")
rel_abun <- read.csv("C:/Users/User/Documents/DREAM/StudyFiles/Training/Training/taxonomy/taxonomy_relabd.species.csv")
```

## Join Data and Select Relevant Columns
Here, the metadata and bacterial abundance data are merged using the `specimen` column as the key. Only the necessary columns from the metadata (specimen, was_term, and NIH.Racial.Category) are selected for the analysis. This ensures that we work with a clean and relevant dataset that combines both the bacterial data and the associated metadata.
```{r}
# Join the datasets, and select only the necessary columns
rel_abun_etn <- metadata %>%
  select(specimen, was_term, NIH.Racial.Category) %>%  
  inner_join(rel_abun, by = "specimen")
```

## Filter Top Bacteria by Variance and Calculate Relative Abundance Percentage
This section filters the bacterial species based on their variance across the samples, selecting the most variable species for further analysis. The user can specify the percentage of variance explained they want to capture, and the code will automatically determine how many bacterial species are needed to reach that percentage. Additionally, the code calculates the percentage of total relative bacterial abundance represented by the selected species and outputs this information. This step is important for focusing the analysis on the most informative bacterial species.
```{r}
# Step 1: Filter variables by variance
bacteria_data <- rel_abun_etn[, 4:ncol(rel_abun_etn)]  # Bacteria data (columns 4 onward)
metadata <- rel_abun_etn[, 1:3]  # Metadata (columns 1 to 3)

# Step 1.1: Calculate total sum of relative abundances for all bacteria
total_abundance <- colSums(bacteria_data)

# Step 1.2: Calculate variance for each bacterium
variance_values <- apply(bacteria_data, 2, var)

# Step 1.3: Sort bacteria by variance in decreasing order
sorted_bacteria_indices <- order(variance_values, decreasing = TRUE)

# Step 1.4: Allow user to define a percentage of explained variance
explained_variance_percentage <- 90  # You can modify this value (as percentage)

# Step 1.5: Calculate cumulative variance to select the number of bacteria that explain the desired variance
cumulative_variance <- cumsum(sort(variance_values, decreasing = TRUE))
total_variance <- sum(variance_values)
selected_bacteria_count <- which(cumulative_variance / total_variance * 100 >= explained_variance_percentage)[1]

# Step 1.6: Filter top bacteria based on selected number (from explained variance)
var_filtered <- bacteria_data[, sorted_bacteria_indices[1:selected_bacteria_count]]

# Step 1.7: Calculate the percentage of total relative abundance represented by the selected bacteria
selected_abundance <- colSums(var_filtered)
percentage_abundance <- sum(selected_abundance) / sum(total_abundance) * 100

# Output the results
cat("Number of selected bacteria:", selected_bacteria_count, "\n")
cat("Percentage of total relative abundance explained by selected bacteria:", round(percentage_abundance, 2), "%\n")
```

## Handle Missing Values in Metadata
In this part, the metadata columns are converted to factors, and the code checks for any missing values in critical variables (NIH.Racial.Category and was_term). If missing values are found, an error message is triggered, halting the process. This ensures that the metadata is complete before proceeding with the analysis, as missing values can disrupt the heatmap and clustering.
```{r error=TRUE}
# Step 2: Check metadata and handle any missing values
metadata$NIH.Racial.Category <- as.factor(metadata$NIH.Racial.Category)
metadata$was_term <- as.factor(metadata$was_term)

if (any(is.na(metadata$NIH.Racial.Category) | is.na(metadata$was_term))) {
  stop("There are missing values in NIH.Racial.Category or was_term")
}
```

## Align Row Names and Prepare Annotation
This section aligns the row names of the filtered bacterial data and metadata to ensure that they match, which is crucial for correctly displaying the data in the heatmap. It also prepares an annotation data frame that will be used to label each sample in the heatmap based on metadata categories like race and term status. The script verifies the levels of these categorical variables, ensuring that they are properly formatted and ready for annotation.
```{r}
# Step 3: Verify that row names are correctly aligned
rownames(var_filtered) <- rownames(metadata) <- 1:nrow(var_filtered)

# Step 4: Prepare the combined annotation for the heatmap
annotation_col <- data.frame(Race = metadata$NIH.Racial.Category,
                             Was_term = metadata$was_term)
rownames(annotation_col) <- rownames(var_filtered)

# Check the levels of the categories
levels(metadata$NIH.Racial.Category)
levels(metadata$was_term)
```

## Define Custom Colors for the Annotations
Here, custom color schemes are defined for the metadata categories, making it easier to visually distinguish between different groups (e.g., different races or term statuses) in the heatmap. The colors for race are selected from a predefined palette, and specific colors are assigned to the two levels of the `was_term` variable (TRUE and FALSE). These colors will be used in the heatmap to annotate the samples.
```{r}
# Step 5: Define colors for both categories
race_colors <- RColorBrewer::brewer.pal(n = length(levels(metadata$NIH.Racial.Category)), 
                                        name = "Set1")
term_colors <- c("FALSE" = "red", "TRUE" = "blue")

# Create the annotation_colors list for pheatmap
annotation_colors <- list(Race = setNames(race_colors, levels(metadata$NIH.Racial.Category)),
                          Was_term = term_colors)
```

## Transpose Bacteria Data and Reorder Specimens
In this step, the bacterial abundance data is transposed so that the bacterial species are represented as rows, and the samples (specimens) are represented as columns. The samples are then reordered based on the `NIH.Racial.Category` variable, which ensures that the samples are grouped by race in the heatmap. This reordering makes it easier to visualize patterns between bacterial abundances and race.
```{r}
# Step 6: Transpose the bacteria data so specimens are columns and bacteria are rows
var_filtered_t <- t(var_filtered)

# Reorder specimens (columns) based on NIH.Racial.Category
ordered_indices <- order(metadata$NIH.Racial.Category)
var_filtered_t_ordered <- var_filtered_t[, ordered_indices]
annotation_col_ordered <- annotation_col[ordered_indices, , drop=FALSE]
```

## Generate and Save Heatmap
This section generates the heatmap using the filtered bacterial data and metadata annotations. The heatmap shows the relative abundances of the selected bacteria across all samples, with annotations for race and term status. The clustering method used ensures that both the bacterial species and the samples are clustered based on similarities. A conditional check is added to handle the number of bacteria. If fewer than 20 bacteria are selected after filtering by variance, their names will be displayed on the heatmap to make it easier to identify them. If there are more than 20 bacteria, the names are hidden to avoid overcrowding the plot and to maintain a cleaner visualization.
The main purpose of this visualization is to explore relationships between bacterial species and metadata categories in a visually intuitive manner.
```{r fig.width=15, fig.height=10, dpi=300}
# Step 7: Generate and save the heatmap with high quality and no scaling
# If the number of selected bacteria is less than or equal to 20, show row names (bacterial names)
if (nrow(var_filtered_t_ordered) <= 20) {
  show_rownames_option <- TRUE  # Show bacterial names
} else {
  show_rownames_option <- FALSE  # Hide bacterial names
}

pheatmap(var_filtered_t_ordered, 
         annotation_col = annotation_col_ordered,  # Ordered annotation for both Race and Was_term
         annotation_colors = annotation_colors,    # Color scheme for NIH.Racial.Category and Was_term
         clustering_method = "complete",           # Cluster only bacteria
         cluster_cols = TRUE,                      # Cluster specimens
         scale = "none",                           # No scaling, keep original relative abundances
         show_rownames = show_rownames_option,     # Conditionally show or hide row names
         show_colnames = FALSE,                    # Hide specimen names
         main = "Heatmap of Bacteria Abundances Grouped by NIH.Racial.Category and Was_term")
```
